Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-07 doc/thesis/abstract.tex
More exposition for ABC; rewrote discussion of real data

We developed a method based on approximate Bayesian computation (ABC) for
estimating structural parameters of the contact network underlying an
observed viral phylogeny. The method combines adaptive sequential Monte Carlo
for ABC, Gillespie simulation for propagating epidemics though networks, and a
{\color{blue}\uline{previously developed}} kernel-based tree similarity score.
{\color{blue}\uline{Our method offers the potential to quantitatively
investigate contact network structure from phylogenies derived from viral
sequence data, as a complementary approach to traditional epidemiological
methods.}}

We applied our method to {\color{red}\sout{fit}}
{\color{blue}\uline{investigate}} the Barab\'{a}si-Albert network model.
{\color{blue}\uline{This model incorporates the mechanism of preferential
attachment observed in real world social and sexual networks, whereby
individuals with a higher number of connections attract new contacts at an
elevated rate (``the rich get richer'').}}
{\color{red}\sout{to simulated transmission trees and applied it to viral
phylogenies estimated from six real-world HIV sequence datasets.}} Using
simulated data, we found that the strength of preferential attachment and the
number of infected nodes in the network could often be accurately estimated. On
the other hand, the mean degree of the network, as well as the total number of
nodes, appeared to be weakly or non-identifiable with ABC.

Finally, the Barab\'{a}si-Albert model was fit to six real world HIV datasets.
We observed substantial heterogeneity in the parameter estimates on real
datasets, with point estimates for the preferential attachment power ranging
from 0.06 to 1.05. 
{\color{blue}\uline{Point estimates of the strength of preferential attachment 
were higher in injection drug user populations, potentially indicating that
high-degree ''hub'' nodes may play a role in epidemics among this risk group.}}
Our results underscore the importance of considering contact structures when
{\color{red}\sout{performing phylodynamic inference}}{\color{blue}\uline{
investigating viral outbreaks.}}

>>>>

We developed a method based on approximate Bayesian computation (ABC) for
estimating structural parameters of the contact network underlying an
observed viral phylogeny. The method combines adaptive sequential Monte Carlo
for ABC, Gillespie simulation for propagating epidemics though networks, and a
{\color{blue}\uline{previously developed}} kernel-based tree similarity score.
{\color{blue}\uline{Our method offers the potential to quantitatively
investigate contact network structure from phylogenies derived from viral
sequence data, complementing traditional epidemiological methods.}}

We applied our method to {\color{red}\sout{fit}}
{\color{blue}\uline{investigate}} the Barab\'{a}si-Albert network model.
{\color{blue}\uline{This model incorporates the preferential attachment
mechanism observed in real world social and sexual networks, whereby
individuals with more connections attract new contacts at an elevated rate
(``the rich get richer'').}}
{\color{red}\sout{to simulated transmission trees and applied it to viral
phylogenies estimated from six real-world HIV sequence datasets.}} Using
simulated data, we found that the strength of preferential attachment and the
number of infected nodes could often be accurately estimated. However, the mean
degree of the network and the total number of nodes appeared to be weakly- or
non-identifiable with ABC.

Finally, the Barab\'{a}si-Albert model was fit to six real world HIV datasets,
and substantial heterogeneity in the parameter estimates was observed.
{\color{red}\sout{Point estimates}}{\color{blue}\uline{Depending on the choice
of prior, posterior means}} for the preferential attachment power ranged from
{\color{red}\sout{0.06 to 1.05}} {\color{blue}\uline{0.26, slightly less than
logarithmic, to 1.00, exactly linear.}} {\color{blue}\uline{Point estimates of
the strength of preferential attachment were higher in two injection drug user
populations, potentially indicating that high-degree ''hub'' nodes may play a
role in epidemics among this risk group.}} Our results underscore the
importance of considering contact structures when {\color{red}\sout{performing
phylodynamic inference}}{\color{blue}\uline{investigating viral outbreaks.}}

### 2016-06-07 doc/thesis/discussion.tex
More exposition for ABC; rewrote discussion of real data

With respect to the preferential attachment power $\alpha$, the six datasets
analysed fell into two categories (\cref{fig:abchpd}). First, we estimated a
preferential attachment power close to 1, indicating linear preferential
attachment, for the BC data and the outbreaks studied by
\textcite{niculescu2015recent} and \textcite{wang2015targeting}. These values
were robust to specifying different priors for $m$. All three datasets were
sampled from populations in which we would expect a high degree of
epidemiological relatedness: \textcite{niculescu2015recent} studied a recent
outbreak among Romanian \gls{IDU}, \citeauthor{wang2015targeting} sampled
acutely infected MSM in Beijing, China, and the BC data constituted a
phylogenetic \gls{IDU} cluster. These are all contexts in which we would expect
some of the assumptions of the BA model, such as a connected network,
relatively high mean degree, and preferential attachment dynamics, to hold.

The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
included in the prior, but these were not robust to changing the prior to
exclude $m = 1$. For the \citeauthor{cuevas2009hiv} data, model
misspecification is likely partially responsible. While the authors found that
a large proportion of the samples were epidemiologically linked, these were
mainly in small local clusters rather than the single large component
postulated by the BA model. In addition, the mixed risk groups in the dataset
would be unlikely to significantly interact, further weakening any global
preferential attachment dynamics. The dataset studied by
\textcite{novitsky2014impact} originated from a densely sampled population
where the predominant risk factor was believed to be heterosexual exposure.
Although the MAP estimate of $\alpha$ was almost unchanged when the value $m =
1$ was excluded from the prior, the confidence interval shrank substantially.

@@

{\color{blue}\uline{
The estimates of the prevalence \gls{I} were largely robust to the choice of
prior on \gls{m}} (\cref{fig:abchpd,fig:abchpdm2}).
\textcite{niculescu2015recent} \uline{reported that 494 new HIV infections were
diagnosed among the studied population (\glspl{IDU} in Bucharest) during the
study period. Our estimate of the prevalence was higher (X), although the 95\%
\gls{HPD} interval did contain the value 494. In addition, it is not
unreasonable to imagine that a substantial fraction of the \gls{HIV}-positive
individuals in an \gls{IDU} population could be undiagnosed, given that these
populations are often marginalized.}

\uline{The individuals sampled by \autocite{wang2015targeting} were recruited by
following a prospective cohort of 2000 \gls{MSM} in Beijing. Of the 2000, 179
new infections were identified during the study period. This number is much
lower than the estimated prevalence of X obtained with \software{netabc}.
However, although} \citeauthor{wang2015targeting} TODO

\textcite{cuevas2009hiv} \uline{reported that 620 newly infected and 1500
chronically infected patients had been sampled during the study period, for a
total of 2120. This is substantially greater than the point estimate of X we
obtained with \gls{ABC}, even without considering the likely possibility that
there were more infected than sampled individuals. Again, this error may in
part be due to model misspecification, which highlights the importance of
selecting a model appropriate to the dataset at hand.}}

>>>>

{\color{red}\sout{With respect to the preferential attachment power $\alpha$,
the six datasets analysed fell into two categories} (\cref{fig:abchpd}).
\sout{First, we estimated a preferential attachment power close to 1,
indicating linear preferential attachment, for the BC data and the outbreaks
studied by} \textcite{niculescu2015recent} and \textcite{wang2015targeting}.
\sout{These values were robust to specifying different priors for $m$. All three
datasets were sampled from populations in which we would expect a high degree
of epidemiological relatedness:} \textcite{niculescu2015recent} \sout{studied a
recent outbreak among Romanian \gls{IDU},} \citeauthor{wang2015targeting}
\sout{sampled acutely infected MSM in Beijing, China, and the BC data
constituted a phylogenetic \gls{IDU} cluster. These are all contexts in which
we would expect some of the assumptions of the BA model, such as a connected
network, relatively high mean degree, and preferential attachment dynamics, to
hold.}

\sout{The remaining three datasets} (\textcite{cuevas2009hiv,
novitsky2014impact, li2015hiv}) \sout{had estimated values of $\alpha$ below
0.5 when $m = 1$ was included in the prior, but these were not robust to
changing the prior to exclude $m = 1$. For the} \citeauthor{cuevas2009hiv} 
\sout{data, model misspecification is likely partially responsible. While the
authors found that a large proportion of the samples were epidemiologically
linked, these were mainly in small local clusters rather than the single large
component postulated by the BA model. In addition, the mixed risk groups in the
dataset would be unlikely to significantly interact, further weakening any
global preferential attachment dynamics. The dataset studied by}
\textcite{novitsky2014impact} \sout{originated from a densely sampled
population where the predominant risk factor was believed to be heterosexual
exposure. Although the MAP estimate of $\alpha$ was almost unchanged when the
value $m = 1$ was excluded from the prior, the confidence interval shrank
substantially.}}

For all datasets we examined, the posterior mean estimates for \gls{alpha} were
sub-linear, ranging from 
    0.27
to 
    0.73.
The sub-linearity is consistent with the results of
\textcite{de2007preferential}, who developed a statistical inference method to
estimate the parameters of a more sophisticated \acrlong{PA} model
incorporating heterogeneous node behaviour. They found \gls{alpha} values
ranging from 0.26 to 0.62, depending on the gender and time period considered. 
Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
this range under both priors, as were the estimates for the
\textcite{wang2015targeting} data and the BC data when \gls{m} = 1 was
disallowed by the prior. The dataset investigated by
\textcite{de2007preferential} was derived from a survey of a random sample of
the Norwegian population, whereas our investigation focused on datasets from
known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
cases. For instance, random sampling is much less likely to discover the
high-degree nodes characterizing the tail of the degree distribution, simply
because those individuals are rare in the general population. In addition, it
is plausible that \gls{HIV}-positive individuals are more likely to be highly
connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
with the number of unprotected sexual contacts.

Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
populations whose primary risk factor for \gls{HIV} infection was heterosexual
contact. \citeauthor{de2007preferential} explicitly excluded reported
homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
heterosexual contact is the primary mode of transmission in Botswana where the
study was done. In the first of the two papers describing the Botswana
study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
was gender-biased, being composed of approximately 75\% women. Our estimate of
\gls{alpha} for these data was X or X, depending on the prior on \gls{m};
\citeauthor{de2007preferential} estimated 0.54, 0.57, and 0.29 for 3-year,
5-year, and lifetime partnership networks respectively for the female portion
of their sample.

For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
populations had a higher estimated \acrlong{PA} power than the other datasets
(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
\textcite{dombrowski2013topological}, who reanalyzed a network of \glspl{IDU}
in Brooklyn, USA, collected between 1991 and
1993~\autocite{friedman2006social}. They found that the the \gls{IDU} network 
resembled a \gls{BA} network much more closely than other social and sexual 
networks, and offered sociological explanations for the apparent \acrlong{PA}
dynamics in this population. Importantly, from a public health perspective,
the authors asserted that the removal of \emph{random} individuals from
\gls{IDU} networks may have the paradoxical effect of increasing the network's
epidemic susceptibility. When low-degree nodes are removed, as would occur
during a police crackdown, their network neighbours may turn to well-known
community members for advice or supplies, thus increasing the connectivity of
these high-degree nodes.

@@

{\color{blue}\uline{
The estimates of the prevalence \gls{I} were largely robust to the choice of
prior on \gls{m}} (\cref{fig:abchpd,fig:abchpdm2}).
\textcite{niculescu2015recent} \uline{reported that 494 new HIV infections were
diagnosed among the studied population (\glspl{IDU} in Bucharest) during the
study period. Our estimate of the prevalence was higher (X), although the 95\%
\gls{HPD} interval did contain the value 494. In addition, it is not
unreasonable that a substantial fraction of the \gls{HIV}-positive individuals
in an \gls{IDU} population could be undiagnosed, given that these populations
are often marginalized.}

\uline{The individuals sampled by} \textcite{wang2015targeting} \uline{were
recruited by following a prospective cohort of 2000 \gls{MSM} in Beijing,
China. Of the 2000, 179 new infections were identified during the study period.
This number is much lower than the estimated prevalence of X obtained with
\software{netabc}, however the authors did not claim to have sampled the entire
sexual network. In a nationwide survey,} \textcite{wu2013hiv} \uline{estimated
that there were 24,198 \gls{MSM} living in Beijing, of whom 5.7\%, or 1379,
were \gls{HIV}-infected, which is within the 95\% confidence interval of our
estimate of \gls{I}.}

\textcite{li2015hiv} \uline{studied an \gls{MSM} population in Shanghai, China.
Despite the apparent similarity to the context of the
}\textcite{wang2015targeting} \uline{study, the estimated \gls{HIV} prevalence
of the } \citeauthor{li2015hiv} \uline{data was higher (X). The aforementioned
survey~\autocite{wu2013hiv} estimated that there were 14,511 \gls{MSM} living
in Shanghai, with an \gls{HIV} prevalence of 6.8\% or approximately 987
individuals. This was within the 95\% confidence interval of the estimate
obtained with \software{netabc}. It is worth noting that} \citeauthor{li2015hiv}
\uline{claimed that there were 80,000 \gls{MSM} living in Shanghai, which is a
significantly different estimate than that obtained by} \textcite{wu2013hiv}.

\textcite{cuevas2009hiv} \uline{reported that 620 newly infected and 1500
chronically infected patients had been sampled during the study period, for a
total of 2120. This is substantially greater than the point estimate of X we
obtained with \gls{ABC}, even without considering the likely possibility that
there were more infected than sampled individuals. Again, this error may in
part be due to model misspecification, which highlights the importance of
selecting a model appropriate to the dataset at hand.} }
